Notes and code examples to accompany "If the Null Fits, You Must Omit

### 2025-11-20 GroupPolarizationStatmod/paper/code_examples.qmd
worked out code example for calculating observed means from latent opinion distributions

---
title: Code examples to accompany "If the Null Fits, You Must Omit"
author: Matt Turner · [matt@mat.phd](mailto:matt@mat.phd)
editor: 
  markdown: 
    wrap: 90
bibliography: this.bib
---

# Purpose: demonstrate gp-stats codebase

@@

## Calculate probability density vector for `Moscovici1969_Americans` condition

See Method section of the paper where we calculate the probabilty density vector for
different measurements based on the latent opinion probability density function.

@@

```{r setup}

```

@@

These

@@

We can plot this simply using just

```{r}
barplot(f_pre, names.arg = -3:3)
```

>>>>

---
title: Notes and code examples to accompany "If the Null Fits, You Must Omit"
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
editor: 
  markdown: 
    wrap: 90
bibliography: this.bib
---

### Purpose: demonstrate gp-stats codebase

@@

# Simulations

I developed two models of group polarization to serve complementary purposes.

## Expectation model of group polarization experiments

See Method section of the paper where we explain how to calculate the probability density
vector representing the probability that a participant gives the opinion represented by
each measurement bin, $b$.

@@

@@

The function `makeProbVec` creates the vector $f[b]$ of probabilities that bin $b_i \in b$
will be reported by a randomly-selected participant, given the measurement bin values and
the distribution of latent opinions, assumed normal, defined by the mean and standard
deviation, $\mu$ and $\sigma$. and bin values. Here bin values are $b=-3,-2,\ldots,3$;
$\mu=-1.2$; and $\sigma=4.3$.

@@

Let's `barplot` to have a look:

```{r out.width=500}
barplot(f_pre, 
        names.arg = -3:3,
        ylim = c(0, 0.4),
        ylab = "Pct. Responding",
        xlab = "Bin Value")
```

Let's check the simulated average observed opinion, $\bar{o}$, using the expected value of
the distribution, i.e., we assume $\bar{o} = \langle o \rangle$, so we use the standard
formula for expected value, which assumes that $f[b]$ is normalized.

$$
\bar{o} = \sum_{b=b_1}^{b_B} b \cdot f[b].
$$

In code,

```{r}
# Calculate simulated and compare with actual
sim_obs_pre <- sum((-3:3) * f_pre)
print(abs(sim_obs_pre - mosc1969_ave_pre))
```

This is within one significant digit greater than the reporting scale, meaning it is an
acceptable fit.

This shows that even just a little spread in the distribution of latent opinions results
in a seriously distorted mean [@Liddell2018]—one that reproduces the mean reported by
@Moscovici1969, but actually confirming the null polarization hypothesis!

Let's check the simulated average observed opinion for post-deliberation, repeating the
previous steps for $t=\mathrm{post}$.

```{r}
f_post <- makeProbVec(-3:3, mu, sd_post)
sprintf("%.2f", f_post)
# Calculate simulated and compare with actual
sim_obs_post <- sum((-3:3) * f_post)
print(abs(sim_obs_post - mosc1969_ave_post))
```

Here we get even closer agreement between the observed post-discussion average opinion in
@Moscovici1969 and the one we simulated.
